### 2016-06-27 publications/Journal/parallel-performance.tex
Parallel section updated
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\paragraph*{Strong and Weak Scaling}
Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size. As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice. The difference between dxex conservative and adevs becoming smaller. The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.\\
In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized. This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue. Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on. Any simulation it performs is immediately void and will have to be reverted as soon as events from kernels 0 and 1 arrive. When kernel 2 reverts, kernel 3 will inevitably have to revert as well since any events it received from kernel 2 are invalidated and marked as such by antimessages. This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
\begin{figure}
	\center

@@

% The really bad
\subsubsection{Interconnect}\label{subsec:parallelinterconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
In interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models. Such a cycle forces sequential operation of the kernels with no speedup possible.

@@

% The okayish
\subsubsection{Phold}
% Same cause, different results
In the Phold model, we first investigate the influence of the percentage of remote events on the speedup. A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
Adevs is significantly slower during conservative synchronization.
Analysis of profiling results shows that exception handling in adevs is the main cause. 
To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling. These functions require the user to implement a state saving mechanism in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel. We feel this would lead to an unfair comparison as we would like to keep the models synchronization-agnostic across all benchmarks.

@@

We slightly modified the Phold benchmark, to include high-priority events.
Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown for Interconnect degrades performance for both optimistic and conservative. This is also the cause of the sublinear speedup observed in our Phold benchmark.

@@

% The real life
\subsubsection{PholdTree}
We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization. We will demonstrate that allocation is critical to achieve a parallel speedup and that the parameters of the model will give either conservative or optimistic an advantage.
\paragraph*{Allocation}\label{PholdTreeallocation}
The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first. 
The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model. 
Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
A single kernel that has an unbalanced number of atomic models or models with unequal computation load in transition functions will slow down the remainder of the chain. This effect is also apparent if the OS unfairly schedules the thread a kernel runs on. In conservative this will lead to excessive polling of the other kernels' eot values, whereas in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel. In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
%
%\begin{figure}
%	\center
%	
%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
%	\caption{PholdTree model breadth first allocation with 3 kernels.}
%	\label{fig:PholdTree_model_bfs}
%	
%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
%	\caption{PholdTree model depth first allocation with 3 kernels.}
%	\label{fig:PholdTree_model_dfs}
%\end{figure}

%Counterintuitively the probability of a high priority event in this model does not affect the performance difference between conservative and optimistic. The key parameter quickly becomes the load of a kernel in models. Our conservative implementation in an uncertain (trivial lookahead) simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.

\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor. The probability of the priority message has no extra impact on performance other than that shown in sequential performance. 
The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization variants. %reword 
The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm. An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails. Even an asynchronous GVT algorithm will involve in a shared memory architecture some form of synchronization, and this will slow down the kernels. \\
Similar to the Queue benchmark the breadth first allocation scheme leads to a chain topology for the kernels, but PholdTree differs in the number of links between the kernels in the chain. In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation. This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree. Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
The performance drop observed for 2 kernels for all allocation schemes and synchronization algorithms is in part due to the higher link count between kernels. With more kernels this effect will be spread across more kernels and result in a relative lower performance penalty. \\
Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels. The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round. As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance. With 4 kernels conservative surpasses optimistic in speedup. This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
\begin{figure}
	\center

@@

\paragraph*{Weak Scaling}
%Intro
In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies. From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models, we want to observe what happens in a closer to linear increase in model count per kernel, so we vary n.

@@

\subsection{Measuring Synchronization Overhead}
The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead. 
Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation. By reducing the cost of both we will isolate the synchronization overhead. We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
	\center

@@

A sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator. Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime. \\

@@

Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation. With each event dynamically allocated this makes it the perfect stress test for any allocation strategy. Both tcmalloc and Boost pools result in an advantage, but not necessarily combined. \\
% Paralell
We next investigate the effect of both strategies in an optimistic simulation. We have already demonstrated that conservative simulation differs little in memory usage from sequential, so we only benchmark the optimistic kernel. From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator. Nonetheless the pool allocator still reduce the allocation overhead. \\

@@

Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS. The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly, useful for example in observing how efficient GVT-triggered deallocation is in practice.\\

@@

A non trivial amount of time in a simulation spent waiting for a random number generator can mask synchronization overhead. Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events. \\

@@

Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. 
The overhead of initially doing so is therefore constant and not a factor in long simulations.\\
Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead. Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a ~ 3 speedup by using the faster random number generator. The parallel kernels have smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor. Dxex conservative is almost insensitive to the changing of the rng. 
Dxex optimistic and adevs conservative are in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.

>>>>

\paragraph*{Strong and Weak Scaling}
Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to dxex sequential for a fixed problem size. As the amount of kernels increases, the optimistic kernel quickly becomes the worst choice. The difference between dxex conservative and adevs becoming smaller. The same effect can be seen for weak scaling in Figure~\ref{fig:Queue_plot_weak}.\\
In Figure \ref{fig:Queue_allocation} the allocation of the Queue model is visualized. This trace allows us to demonstrate the benchmark results and explain why optimistic is not ideal for such a chain topology as present in Queue. Kernel 2 is dependent on events from kernels 0 and 1 but by definition of the optimistic synchronization protocol it will simulate ahead without waiting for events it depends on. Any simulation it performs will have to be reverted as soon as events from kernels 0 and 1 arrive. When kernel 2 reverts, kernel 3 will inevitably have to revert as well since most of the events it received from kernel 2 are invalidated and marked as such by antimessages. This leads to severe performance degradation, with an increasing probability of cascading reverts as the number of kernels and models per kernels increases.
The speedup of adevs is always in comparison with the runtime of the corresponding dxex sequential benchmark.
\begin{figure}
	\center

@@

% The really bad
\subsubsection{Interconnect}\label{subsec:parallelinterconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and an increasing number of of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.
In Interconnect there is no allocation scheme possible that avoids cyclic dependencies between simulation kernels, as shown in the trace \ref{fig:interconnect_allocation_parallel} of a simulation with 4 models. Such a cycle forces sequential operation of the kernels with no speedup possible.

@@

% The okayish
\subsubsection{Phold}
% Same cause, different results
In the Phold model, we first investigate the influence of the percentage of remote events on the speedup. A remote event in this context is an event that is sent from a model on one kernel to a model on another simulation kernel.
When remote events are rare, optimistic synchronization rarely has to roll back, thus increasing performance.
With more common remote events, however, optimistic synchronization quickly slows down due to frequent rollbacks.
Conservative synchronization, on the other hand, is mostly unconcerned with the number of remote events: the mere fact that a remote event can happen, causes it to block and wait.
Even though a single synchronization protocol is always ideal in this case, it already shows that different synchronization protocols respond differently to a changing model.
Adevs is significantly slower during conservative synchronization.
Analysis of profiling callgraphs shows that exception handling in adevs is the main cause. 
To keep the models equivalent, the adevs version does not provide the \{begin,end\}Lookahead methods, which accounts for the exception handling. These functions require the user to implement a state saving in contrast to PythonPDEVS and dxex's optimistic kernels which handle this inside the kernel. We feel this would lead to an unfair comparison as we would like to keep the models synchronization-agnostic across all benchmarks.

@@

We slightly modified the Phold benchmark, to include high-priority events.
Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.
In Figure \ref{fig:phold_allocation} it is clear that in Phold it is possible for dependency cycles to form between kernels which as we have shown in Interconnect degrades performance for both optimistic and conservative. This is also the cause of the sublinear speedup observed in our Phold benchmark.

@@

% The real life
\subsubsection{PholdTree}
We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization. We will demonstrate that allocation is critical to achieve a parallel speedup and that the configuration of the model will give either conservative or optimistic an advantage.
\paragraph*{Allocation}\label{PholdTreeallocation}
The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first. 
The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model. 
Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
A single kernel that has an unbalanced number of atomic models or unequal computation load in transition functions will slow down the remainder of the chain. This effect is also apparent if the thread a kernel runs on is not fairly scheduled. In conservative this will lead to excessive polling of the other kernels' eot values, in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel. In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.
%
%\begin{figure}
%	\center
%	
%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeBFS.eps}
%	\caption{PholdTree model breadth first allocation with 3 kernels.}
%	\label{fig:PholdTree_model_bfs}
%	
%	\includegraphics[width=\modelfraction\columnwidth]{fig/pholdtreeDFS.eps}
%	\caption{PholdTree model depth first allocation with 3 kernels.}
%	\label{fig:PholdTree_model_dfs}
%\end{figure}

\paragraph*{Strong Scaling}\label{pholdtreestrongscale}
In Figure \ref{fig:PholdTree_plot_alloc_high} we see that the difference in performance for all 4 kernel configurations compared to that shown in Figure \ref{fig:PholdTree_plot_alloc_low} is a constant factor. The probability of the priority message has no extra impact on performance other than that shown in sequential performance.

The probability of a high priority event in this model does not affect the performance difference between conservative and optimistic. The key parameter quickly becomes the load of a kernel in models. Our conservative implementation in an uncertain simulation is very sensitive to a high load in models, whereas optimistic has no such limitation.

The difference between depth first and breadth first allocated kernels is striking, the first results in sublinear speedup for both synchronization protocols. %reword 
The breadth first allocation scheme will lead to a very high number of inter-kernel connections, which is detrimental for any parallel synchronization algorithm. An event exchanged between kernels cannot be securely deallocated without a GVT algorithm and all the complexity this entails. Even a fast asynchronous GVT algorithm will require some form of inter kernel synchronization and span a timeframe during which allocated memory cannot be reused, forcing new allocations.
Similar to the Queue benchmark the breadth first allocation scheme leads to a topology for the kernels resembling a chain but with more branches in the chain. In Queue there is only a single model on the edge of a kernel exchanging messages to a single other model in a neighbouring kernel, this is not the case in the PholdTree under breadth first allocation. This explains the difference in speedup between Queue and PholdTree despite the similarity in kernel topology.\\
Depth first allocation still has a higher inter kernel connection count, but not on the same order as breadth first allocated PholdTree. Depth first allocation will converge to a star topology which in this benchmark leads to a significant speedup.\\
The performance drop observed for 2 kernels for all allocation schemes and synchronization protocols is in part due to the higher link count between kernels. With more kernels these connections will be spread across more kernels and result in a relative lower performance penalty. \\
Conservative synchronization suffers a further penalty in this benchmark when there are only a few kernels. The PholdTree model has a lookahead of $\epsilon$ leading the kernel to query each allocated atomic model for a lookahead value on each simulation round. As the atomic models are spread over more kernels this effect is reduced leading to an increase in performance. With 4 kernels conservative surpasses optimistic in speedup. This is clear evidence that conservative synchronization is a good candidate for parallel simulation even in simulations with uncertainty.
\begin{figure}
	\center

@@

\paragraph*{Weak Scaling}
%Intro
In this paragraph we investigate dxex's parallel performance when the the number of atomic models distributed over a fixed number of kernels varies. From \ref{eq:pholdtreemodelcount} we know that increasing d will lead to a very rapid increase in models. We want to observe what happens in a closer to linear increase in model count per kernel by varying n.

@@

\subsection{Measuring Synchronization Overhead}
The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead.

Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation. By reducing the cost of both we will isolate the synchronization overhead. We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
	\center

@@

In a sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator. Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime. \\

@@

Interconnect is a perfect test case since it generates each round a set of events quadratic in size to the number of atomic models in the simulation. With each event dynamically allocated, this makes it the perfect stress test for any allocation strategy. Both tcmalloc and Boost pools result in an advantage, but not necessarily combined. \\
% Parallel
We next investigate the effect of both strategies in an optimistic simulation. We have already demonstrated that conservative simulation differs little in memory usage from sequential, so we only benchmark the optimistic kernel. From the results it is clear that optimistic simulation benefits greatly from the use of tcmalloc, regardless of the allocator. Nonetheless the pool allocator still reduces the allocation overhead. \\

@@

Both the pools and tcmalloc will try to keep memory allocated to them by the OS, this gives a somewhat misleading view in statistics collected from the OS. The dxex kernels themselves and the pool interfaces allow the user to instrument event allocation and deallocation exactly which can be useful in measuring how efficient GVT-triggered deallocation is in practice.\\

@@

A non trivial amount of time in a simulation spent waiting for a random number generator (rng) can mask synchronization overhead. Several of our benchmarks use random number generators to test the effect of uncertainty in a simulation, either by random time advances or random destinations for events. \\

@@

Dxex's conservative kernel is insensitive to storing the rng object in the atomic model state, since no copying/state saving occurs in dxex conservative simulation. \\
Replacing the STL rng with the threading optimized trng we are more accurately able to isolate synchronization overhead. Figure \ref{fig:Queuerngspeedup} shows that both dxex and adevs in sequential simulation gain a factor ~3 speedup by using the faster random number generator. The parallel kernels have a smaller speedup with adevs' conservative and dxex optimistic achieving an almost identical factor. Dxex conservative is almost insensitive to the changing of the rng. 
Dxex optimistic and adevs conservative in the default configuration spend relatively more time waiting for random numbers than dxex conservative, demonstrated by the speedup gained when random number generation is accelerated.
